\bf Title here

### 2020-08-26 paper/paper.tex
update .pdf &.tex

\bigskip
\begin{abstract}
Friedman \& Tukey commented on their initial paper on projection pursuit
in 1974 that ``the technique used for maximising the projection index
strongly influences both the statistical and the computational aspects
of the procedure.'' While many projection pursuit indices have been
proposed in the literature, few concerns the optimisation procedure. In
this paper, we developed a system of diagnostics aiming to visually
learn how the optimisation procedures find its way towards the optimum.
This diagnostic system can be applied to more generally to help
practitioner to unveil the black-box in randomised iterative
(optimisation) algorithms. An R package, ferrn, has been created to
implement this diagnostic system.
\end{abstract}

@@

Given an optimisation problem, two basic approaches find the optimum
based on different thinking. An analytical approach aims to find the
optimal solution in a finite number of steps, but a potential issue with
it is that the close-form solution may not be available when the problem
starts to become complex. An iterative approach, on the other hand,
finds the optimum based on the idea of making progressive improvement to
the current solution. An iterative method may end up finding a local
optimum but the progressive nature of the algorithm allows the
practitioner to decide when to stop if a desirable accuracy has been
achieved.

A traditional while often used in practice is an iteartive method called
\emph{line search method} \citep{fletcher2013practical}. In a simple
one-dimensional problem of finding the \(x\) that minimises \(f(x)\),
line search achieves the goal via an iterative algorithm in the form of
Equation \ref{eq:line-search}.

@@

Modern development of the line search methods focuses on proposing
different computation on the seraching direction: \(d^{(j)}\) and
various approximations on the step size: \(\alpha_j\) catered for
practical optimisation problems. The specific problem context we are
interested in is called projection pursuit guided tour. Projection
pursuit and guided tour are two separate methods in exporatory data
analysis focusing on different aspects: coined by
\citet{friedman1974projection}, projection pursuit detects interesting
structures (i.e.~clustering, outliers and skewness) in multivariate data
via low dimensions projection; whilst guided tour is a particular
variation in a broader class of data visualisation method called tour.

@@

Before introducing the guided tour, we shall be familiar with the
general tour method \citep{cook2008grand}. A tour produces animated
visualisation of the high dimensional data via rotating low dimension
planes. The smoothness of the animation is ensured by computing a series
of intermediate planes between two low dimension planes via geodesic
interpolation and we refer readers to \citet{buja2005computational} for
the mathematical details. Iteratively choose different low dimension
planes and interpolate between them forms a tour path. Different types
of tour methods choose the low dimensional planes differently and we
mention two other type of tour that are commonly used. A grand tour
selects the planes randomly in the high dimensional space and hence
serves as an initial exploration of the data. Manual control allows
researches to fine-tuning an existing projection by gradually phase in
and out one variable.

@@

where \(I_d\) is the d-dimentional identity matrix. The constraint
requires the projection basis \(\mathbf{A}\) to be a orthonogal matrix
with each column vector being orthornormal.

There are several features of this optimisation that are worth noticing.
First of all, it is a multivariate constraint optimisation problem.
Since the decision variables are the entries of a projection basis, it
is required to be orthonormal. It is also likely that the objective
function is non-differentiable or the gradient information is simply not
available. In this case, we will need to either use some approximation
of the gradient or turn to derivative free methods. Given the goal of
projection pursuit as finding the basis with the largest index value,
the optimisation problem need to be able to find the global maximum.
Along the way, locla maximum may also be of our interest since they
could present unexpected interesting projections. There is also one
computational consideration: the optimisation procedure needs to be easy
to compute since the tour animation is played in real-time.

@@

Below we introduce three possible algorithms: \texttt{search\_better}
and \texttt{search\_better\_random} are derivative free methods that
sample candidate basis in the neighbourhood whilst
\texttt{search\_geodesic} approximates the gradient information and acts
as a non-derivative version of the gradient acsent.

\citet{posse1995projection} proposed a random search algorithm that
samples a candidate basis \(\mathbf{A}_{l}\) in the neighbourhood of the
current basis \(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}\),
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{\text{rand}}\) is a randomly generated matrix with the
same dimension as \(\mathbf{A}_{\text{cur}}\). The optimiser keeps
sampling bases near the current basis until it finds one with higher
index value than the current basis and then outputs it for guided tour
to construct the interpolation path. A new round of search continues to
find a better basis after the interpolation finishes. The halving
parameter \(c\) with default value of 30 is designed to adjust the
searching neighbourhood \(\alpha\). When the search needs to sample more
than \(c\) number of basis to find an accpeted basis, the neighbourhood
parameter \(\alpha\) will be reduced by half in the next iteration. If
the optimiser can't find a better basis within the maximum number of
tries \(l_{\max}\), the algorithm stops. The algorithm is summarised in
Algorithm \ref{random-search} for one iteration. {[}mention
orthonormalise to ensure the constaint is fulfilled; don't use
derivative information but a random search{]}

@@

Simulated annealing
\citep[\citet{kirkpatrick1983optimization}]{bertsimas1993simulated}
modifies \texttt{search\_better} based on a non-increasing cooling
scheme \(T(j)\). Given an initial \(T_0\), the temperature at iteration
\(k\) is defined as \(T(j) = \frac{T_0}{\log(j + 1)}\). When a candidate
basis fails to have an index value larger than the current basis,
simulated annealing gives it a second chance to be accepted with
probabiliby
\[P= \min\left\{\exp\left[-\frac{I_{cur} - I_{j}}{T(j)}\right],1\right\}\]
where \(I_{(\cdot)}\) denotes the index value of a given basis. This
implementation allows the algorithm to jump out of a local maximum and
enables a more holistic search of the whole parameter space. This
feature is particularly useful when the dimension of the projected space
is smaller than the number of informative variables in the dataset
(i.e.~a one dimenional projection of the dataset with two informative
variables). The algorithm can be written as replacing line 5-11 of
Algorithm \ref{random-search} with Algorithm \ref{simulated_annealing}.

@@

\citet{cook1995grand} used a gradient ascent algorithm with
pseudo-derivative. Instead of computing the actual gradient of the index
function with respect to the projection basis matrix,
\texttt{search\_geodesic} samples \(2n\) bases that are randomly
generated on a uniform open ball with the radius controlled by the
\(\delta\) parameter and finds the most promising one to construct the
geodesic direction as an approximation to the graident. The \(\delta\)
parameter is usually set to be tiny to ensure the geogesic direction is
a good local approximation to the graident. Once the searching direction
is determined, the optimiser finds the best projection
\(\mathbf{A}_{**}\) on the geodesic as the candidate.
\(\mathbf{A}_{**}\) is outputted for the current iteration if the
percentage change in the index value between \(\mathbf{A}_{**}\) and
\(\mathbf{A}_{\text{cur}}\) is greater than a threshold value or the
algorithm repeats the above steps until \(l_{\max}\) is reached and the
searching terminates. Algorithm \ref{search-geodesic} summarise the
steps in geodesic search.

@@

It is worth-noticing that the data structure constructed above meets the
tidy data principal \citep{wickham2014tidy} that states

@@

A primary interest of diagnosing an optimisation algorithm is to study
how it progressively finds its optimum. One way of doing it is to plot
the index value across its natural ordering \(t\), however, it usually
takes the algorithm much longer to find a better basis than the current
one towards the end of the search. When plotting the searching
observations as points in a plot, the space each iteartion take will be
proportional to the number of points in that iteration.

Another option is to use summarisation for each iteration. Boxplot is a
suitable candidate that provides five points summary of the data,
however, there are two pieces of information missing from the boxplot:
1) It does not report the number of points, and 2) the position of the
last basis. This could be remedised by adding more layers using the
concept of gramma of graphics. A label geometry is added at the bottom
of the plot to show the number of points in each iteration and a line
geometry links the last basis in each iteration. Further, a option to
switch betwen displaying points and boxplot geometry is helpful since
point geometry can be more intuitive for the iteration with few
observations. This is avhieved via a \texttt{cutoff} parameter. Below we
define the searching point plot based on four different component layers

@@

  \begin{itemize}
  \tightlist
  \item
    data: group by \(j\) and filter the observations in the group that
    have count greater than \texttt{cutoff\ =\ 15}.
  \item
    x: \(j\) is mapped to the x-axis
  \item
    y: the statistical transformed index value: \(Q_{I^{\prime}_t}(q)\)
    is mapped to the y-axis where \(Q_X(q)\), \(q = 0, 25, 50, 75, 100\)
    finds the qth-quantile of \(X\) and \(I^{\prime}_t\) denotes the
    index value of all the searching bases defined in Matrix
    \ref{eq:data-structure}.
  \end{itemize}
\item
  Layer 2: point geom

@@

Figure \ref{toy-search} presents a comparison between the two plotting
options discussed above. Using the natural order to plot searching
points distorts the point from the first few iterations and
over-emphasize the searches in the last few iterations. The second plot
design evenly presents the summarised information for each iteration
while allow for a switch to the full information for the iteration with
small number of observations.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/toy-search-1.pdf}
\caption{\label{toy-search}A comparison of plotting the same search
points with different plot designs. In upper plot, points from the last
three iterations span the vast majority of the plotting space leaving
the first few iterations being squeezed together whilst the lower plot
spaces each iteration evenly and presents summary information easy to
read.}
\end{figure}

@@

Figure \ref{toy-interp} presents the interpolation of three different
tour paths. The upper plot shows a desirable interpolating in each
iteration with the index value being progressively and monotonically
increasing. While in the middle plot, the increases towards the target
basis is not monotonical in the last two iterations and interpolated
basis with higher index value can be found on the tour path. The lower
plot is constructed using \texttt{search\_better\_random}, where a basis
with smaller index value has a probabilistic chance of being accepted
and we can observe a much more involved pattern. In iteration three, the
probablistic acceptance produces a monotonically decreasing
interpolation whilst in iteration five, six and seven, with also a
probablistic acceptance, the interpolation is now non-monotical. When
the target basis has a higher index value, iteration eight reaches this
basis by first decreases the index value. While the interpolation plot
shows different possibilities of the change in index value on the tour
path, the diagnose of the validity of each pattern and the related
optimisation algorithms will be postponed to Section \ref{application}.

>>>>

\bigskip
\begin{abstract}
Friedman \& Tukey commented on their initial paper on projection pursuit
in 1974 that ``the technique used for maximising the projection index
strongly influences both the statistical and the computational aspects
of the procedure.'' While many projection pursuit indices have been
proposed in the literature, few concerns the optimisation procedure. In
this paper, we developed a system of diagnostics aiming to visually
learn how the optimisation procedures find its way towards the optimum.
This diagnostic system can be applied more generally to help
practitioner to unveil the black-box in randomised iterative
(optimisation) algorithms. An R package, ferrn, has been created to
implement this diagnostic system.
\end{abstract}

@@

Given an optimisation problem, two basic approaches find the optimum
based on different thinking. An analytical approach aims to find the
optimal solution in a finite number of steps, but a potential issue with
it is that the closed-form solution may not be available when the
problem starts to become complex. An iterative approach, on the other
hand, finds the optimum based on the idea of making progressive
improvement to the current solution. An iterative method may end up
finding a local optimum but the progressive nature of the algorithm
allows the practitioner to decide when to stop if a desirable accuracy
has been achieved.

A traditional while often used in practice is an iterative method called
\emph{line search method} \citep{fletcher2013practical}. In a simple
one-dimensional problem of finding the \(x\) that minimises \(f(x)\),
line search achieves the goal via an iterative algorithm in the form of
Equation \ref{eq:line-search}.

@@

Modern development of the line search methods focuses on proposing
different computation on the searching direction: \(d^{(j)}\) and
various approximations on the step size: \(\alpha_j\) catered for
practical optimisation problems. The specific problem context we are
interested in is called projection pursuit guided tour. Projection
pursuit and guided tour are two separate methods in exporatory data
analysis focusing on different aspects: coined by
\citet{friedman1974projection}, projection pursuit detects interesting
structures (i.e.~clustering, outliers and skewness) in multivariate data
via low dimensions projection; whilst guided tour is a particular
variation in a broader class of data visualisation method called tour.

@@

Before introducing the guided tour, we shall be familiar with the
general tour method \citep{cook2008grand}. A tour produces animated
visualisation of the high dimensional data via rotating low dimension
planes. The smoothness of the animation is ensured by computing a series
of intermediate planes between two low dimension planes via geodesic
interpolation and we refer readers to \citet{buja2005computational} for
the mathematical details. Iteratively choosing different low dimension
planes and interpolating between them forms a tour path. Different types
of tour methods choose the low dimensional planes differently and we
mention two other types of tour that are commonly used. A grand tour
selects the planes randomly in the high dimensional space and hence
serves as an initial exploration of the data. Manual control allows
researches to fine-tuning an existing projection by gradually phase in
and out one variable.

@@

where \(I_d\) is the d-dimensional identity matrix. The constraint
requires the projection basis \(\mathbf{A}\) to be an orthogonal matrix
with each column vector being orthonormal.

There are several features of this optimisation that are worth noticing.
First of all, it is a multivariate constraint optimisation problem.
Since the decision variables are the entries of a projection basis, it
is required to be orthonormal. It is also likely that the objective
function is non-differentiable or the gradient information is simply not
available. In this case, we will need to either use some approximation
of the gradient or turn to derivative free methods. Given the goal of
projection pursuit as finding the basis with the largest index value,
the optimisation problem needs to be able to find the global maximum.
Along the way, local maximum may also be of our interest since they
could present unexpected interesting projections. There is also one
computational consideration: the optimisation procedure needs to be easy
to compute since the tour animation is played in real-time.

@@

Below we introduce three possible algorithms: \texttt{search\_better}
and \texttt{search\_better\_random} are derivative free methods that
sample candidate basis in the neighbourhood whilst
\texttt{search\_geodesic} approximates the gradient information and acts
as a non-derivative version of the gradient ascent.

\citet{posse1995projection} proposed a random search algorithm that
samples a candidate basis \(\mathbf{A}_{l}\) in the neighbourhood of the
current basis \(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}\),
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{\text{rand}}\) is a randomly generated matrix with the
same dimension as \(\mathbf{A}_{\text{cur}}\). The optimiser keeps
sampling bases near the current basis until it finds one with higher
index value than the current basis and then outputs it for guided tour
to construct the interpolation path. A new round of search continues to
find a better basis after the interpolation finishes. The halving
parameter \(c\) with default value of 30 is designed to adjust the
searching neighbourhood \(\alpha\). When the search needs to sample more
than \(c\) number of basis to find an accepted basis, the neighbourhood
parameter \(\alpha\) will be reduced by half in the next iteration. If
the optimiser can't find a better basis within the maximum number of
tries \(l_{\max}\), the algorithm stops. The algorithm is summarised in
Algorithm \ref{random-search} for one iteration. {[}mention
orthonormalise to ensure the constraint is fulfilled; don't use
derivative information but a random search{]}

@@

Simulated annealing
\citep[\citet{bertsimas1993simulated}]{kirkpatrick1983optimization}
modifies \texttt{search\_better} based on a non-increasing cooling
scheme \(T(j)\). Given an initial \(T_0\), the temperature at iteration
\(k\) is defined as \(T(j) = \frac{T_0}{\log(j + 1)}\). When a candidate
basis fails to have an index value larger than the current basis,
simulated annealing gives it a second chance to be accepted with
probability
\[P= \min\left\{\exp\left[-\frac{I_{\text{cur}} - I_{j}}{T(j)}\right],1\right\}\]
where \(I_{(\cdot)}\) denotes the index value of a given basis. This
implementation allows the algorithm to jump out of a local maximum and
enables a more holistic search of the whole parameter space. This
feature is particularly useful when the dimension of the projected space
is smaller than the number of informative variables in the dataset
(i.e.~a one dimensional projection of the dataset with two informative
variables). The algorithm can be written as replacing line 5-11 of
Algorithm \ref{random-search} with Algorithm \ref{simulated_annealing}.

@@

\citet{cook1995grand} used a gradient ascent algorithm with
pseudo-derivative. Instead of computing the actual gradient of the index
function with respect to the projection basis matrix,
\texttt{search\_geodesic} samples \(2n\) bases that are randomly
generated on a uniform open ball with the radius controlled by the
\(\delta\) parameter and finds the most promising one to construct the
geodesic direction as an approximation to the gradient. The \(\delta\)
parameter is usually set to be tiny to ensure the geodesic direction is
a good local approximation to the gradient. Once the searching direction
is determined, the optimiser finds the best projection
\(\mathbf{A}_{**}\) on the geodesic as the candidate.
\(\mathbf{A}_{**}\) is outputted for the current iteration if the
percentage change in the index value between \(\mathbf{A}_{**}\) and
\(\mathbf{A}_{\text{cur}}\) is greater than a threshold value or the
algorithm repeats the above steps until \(l_{\max}\) is reached and the
searching terminates. Algorithm \ref{search-geodesic} summarises the
steps in geodesic search.

@@

It is worth noticing that the data structure constructed above meets the
tidy data principle \citep{wickham2014tidy} that states

@@

A primary interest of diagnosing an optimisation algorithm is to study
how it progressively finds its optimum. One way of doing it is to plot
the index value across its natural ordering \(t\), however, it usually
takes the algorithm much longer to find a better basis than the current
one towards the end of the search. When plotting the searching
observations as points in a plot, the space each iteration takes will be
proportional to the number of points in that iteration.

Another option is to use summarisation for each iteration. Boxplot is a
suitable candidate that provides five points summary of the data,
however, there are two pieces of information missing from the boxplot:
1) It does not report the number of points, and 2) the position of the
last basis. This could be remedied by adding more layers using the
concept of grammar of graphics. A label geometry is added at the bottom
of the plot to show the number of points in each iteration and a line
geometry links the last basis in each iteration. Further, an option to
switch between displaying points and boxplot geometry is helpful since
point geometry can be more intuitive for the iteration with few
observations. This is achieved via a \texttt{cutoff} parameter. Below we
define the searching point plot based on four different component layers

@@

  \begin{itemize}
  \tightlist
  \item
    data: group by \(j\) and filter the observations in the groups that
    have count greater than \texttt{cutoff\ =\ 15}.
  \item
    x: \(j\) is mapped to the x-axis
  \item
    y: the statistical transformed index value: \(Q_{I^{\prime}_t}(q)\)
    is mapped to the y-axis where \(Q_X(q)\), \(q = 0, 25, 50, 75, 100\)
    finds the qth-quantile of \(X\) and \(I^{\prime}_t\) denotes the
    index value of all the searching bases defined in Matrix
    \ref{eq:data-structure}.
  \end{itemize}
\item
  Layer 2: point geom

@@

Figure \ref{toy-search} presents a comparison between the two plotting
options discussed above. Using the natural order to plot searching
points distorts the point from the first few iterations and
over-emphasizes the searches in the last few iterations. The second plot
design evenly presents the summarised information for each iteration
while allowing for a switch to the full information for the iteration
with small number of observations.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/toy-search-1.pdf}
\caption{\label{toy-search}A comparison of plotting the same search
points with different plot designs. In the upper plot, points from the
last three iterations span the vast majority of the plotting space
leaving the first few iterations being squeezed together whilst the
lower plot spaces each iteration evenly and presents summary information
easy to read.}
\end{figure}

@@

Figure \ref{toy-interp} presents the interpolation of three different
tour paths. The upper plot shows a desirable interpolation in each
iteration with the index value being progressively and monotonically
increasing. While in the middle plot, the increases towards the target
basis is not monotonical in the last two iterations and interpolated
basis with higher index value can be found on the tour path. The lower
plot is constructed using \texttt{search\_better\_random}, where a basis
with smaller index value has a probabilistic chance of being accepted
and we can observe a much more involved pattern. In iteration three, the
probabilistic acceptance produces a monotonically decreasing
interpolation whilst in iteration five, six and seven, with also a
probabilistic acceptance, the interpolation is now non-monotonical. When
the target basis has a higher index value, iteration eight reaches this
basis by first decreasing the index value. While the interpolation plot
shows different possibilities of the change in index value on the tour
path, the diagnosis of the validity of each pattern and the related
optimisation algorithms will be postponed to Section \ref{application}.
